Improvement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-06-08 rigid_bodies.tex
Revision of backward error analysis and NVT Adjustment

For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
Here, we prove that their findings can be extended to the symplectic integration of rigid-body motion.

@@

The first new result we present in this paper is an analytical form of an approximation to the shadow Hamiltonian, referred here as \textit{refine}d Hamiltonian, concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~ \citenum{Silveira_2017}.
A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian} and here we summarize the final expressions.

As demonstrated in Appendix~\ref{sec:rigid body shadow hamiltonian}, the \textit{refined} Hamiltonian for a Verlet-type method applied to a system of rigid bodies is given by $\refined{\Ham{}} = \refined K + \refined U + \mathcal{O}(h^4)$, where $\refined K$ and $\refined U$ are \textit{refined} kinetic and potential energies.
The latter is given by
\begin{equation}
\label{eq:modified potential energy}
\refined U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
\end{equation}
Note that $\refined U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
In turn, the \textit{refined} kinetic energy can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\refined K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\refined{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, given by $\tilde{\mt \Omega} = {\mt \Omega} + {\timestep^2}/{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)$, where
\begin{equation*}
{\mt \Omega} = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right]
\end{equation*}
and ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, given by
\begin{equation*}
{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
U_{\vt r \vt r} & U_{\vt r \vt q} \\
U_{\vt q \vt r} & U_{\vt q \vt q}
\end{array}\right].
\end{equation*}

Hence, the trajectory generated by the splitting scheme of Eq.~\ref{eq:splitting NVE}, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
\begin{equation*}
\dot{\vt r} = \diff{\refined K}{\vt p}, \;
\dot{\vt q} = \diff{\refined K}{\vt \pi}, \;
\dot{\vt p} = -\diff{\refined{\Ham{}}}{\vt r}, \; \text{and} \;
\dot{\vt \pi} = -\diff{\refined{\Ham{}}}{\vt q}.
\end{equation*}

@@

Analytical calculation of $\refined U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
\end{equation}

@@

In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each rigid body $i$, such as required for satisfying the unit-norm condition \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of\cite{Schay_1995}
\begin{equation*}
\dot{\vt q}^{[3]}_t = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
\end{equation*}
where ${\mt \Pi}$ is a matrix-valued function given by either product ${\mt B}\tr{\mt B}$ or ${\mt C}\tr{\mt C}$ \cite{Silveira_2017}.
In practice, we evaluate the equation above individually for each body $i$ with ${\mt \Pi}({\vt q}_i) = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix, thus projecting the interpolation-based derivative quaternion into the hyperplane orthogonal to ${\vt q}_i$.

\subsection{Adjustment of the NVT Dynamics}
\label{sec:refined_method}
Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting scheme, as described before.
The only required modification is that $\refined K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that $G_1$ becomes
\begin{equation}
\label{eq:thermostat_force}
G_1 = 2 \refined K - L k_\text{\tiny B} T.
\end{equation}
Also, the ``extended energy'' preserved by this approach is now given by
\begin{equation}
\label{eq:this extended energy}
\widetilde{H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

One instant in which $\refined K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
Fortunately, only positions and orientations are required for the numerical differentiation.
This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time, as schematically illustrated in Fig.~\ref{fig:virtual}.
By incomplete we mean that the step is aborted once the positions and orientations have been updated.
\begin{figure}
\begin{tikzpicture}
\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
\node at (0.0,-0.2)[below]{$r_{t-h}$};
\node at (2.2,-0.2)[below]{$r_{t}$};
\node at (4.4,-0.2)[below]{$r_{t+h}$};
\node at (6.6,-0.2)[below]{$r_{t+2h}$};
\path[->,thick] (1) edge [bend left=45] (2);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
\path[dashed,->,thick] (1) edge [bend right=45] (0);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
\path[dashed,->,thick] (2) edge [bend left=45] (3);
\end{tikzpicture}
\caption{Schematic representation}
\label{fig:virtual}
\end{figure}

@@

\appendix

\section{Backward Error Analysis}
\label{sec:rigid body shadow hamiltonian}

@@

As will become clear below, it is convenient to rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form
\begin{equation*}
K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} {\mt \Omega}(\vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation*}
where $\mt \Omega$ is a symmetric, block-diagonal matrix of size $7N \times 7N$ defined as
\begin{equation*}
{\mt \Omega} = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right].
\end{equation*}

>>>>

For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
Here, we prove that their findings can be extended to the symplectic integration of rigid-body motion.

@@

A new result we present here is an analytically derived approximation for $\Ham{S}$, referred to here as a \textit{refined Hamiltonian} $\refined{\Ham{}}$, concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~\citenum{Silveira_2017}.
A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian}.
As demonstrated therein, such approximation is given by $\Ham{S} = \refined K + \refined U + \mathcal{O}(h^4)$, where $\refined K$ and $\refined U$ are the refined versions of the kinetic and potential energies, respectively.
The latter is given by
\begin{equation}
\label{eq:modified potential energy}
\refined U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
\end{equation}

Note that $\refined U$ becomes dependent on $\timestep^2$, but keeps being independent of $\vt p$ and $\vt \pi$.
In turn, the \textit{refined kinetic energy} can be expressed as
\begin{equation}
\label{eq:modified kinetic energy}
\refined K = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega}$.
In this definition, ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to $\vt r$ and $\vt q$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
Although $\refined{\mathbf \Omega}$ is symmetric like $\mt \Omega$, its structure is not block-diagonal due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both $\vt r$ and $\vt q$.
As a result, one cannot generally split the refined kinetic energy into independent translational and rotational contributions.

Now notice that a trajectory generated by the splitting scheme of Eq.~\eqref{eq:splitting NVE}, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will even more closely approach the solution of a modified ODE system given by
\begin{equation*}
\dot{\vt r} = \diff{\refined K}{\vt p}, \;
\dot{\vt q} = \diff{\refined K}{\vt \pi}, \;
\dot{\vt p} = -\diff{\refined{\Ham{}}}{\vt r}, \; \text{and} \;
\dot{\vt \pi} = -\diff{\refined{\Ham{}}}{\vt q}.
\end{equation*}

@@

Analytical calculation of $\refined U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
Fortunately, we can estimate it rather easily by following \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, who employed numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
\begin{equation}
\label{eq:modified kinetic energy estimator}
\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
\end{equation}

@@

In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each body $i$, such as required for preserving the unit-norm constraint \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of \cite{Schay_1995}
\begin{equation*}
\dot{\vt q}^{[3]}_t = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
\end{equation*}
where ${\mt \Pi}(\vt q)$ is a block-diagonal projection matrix, with each block given by ${\mt \Pi}_i = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix.

\subsection{Adjustment of the NVT Dynamics}
\label{sec:refined_method}

Bringing the attention back to the NVT case, the observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a canonical ensemble for a system with this Hamiltonian.
Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $6N-3$ degrees of freedom is that
\begin{equation*}
\left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T,
\end{equation*}
which becomes $2\langle \refined{K} \rangle = (6N - 3) k_\text{\tiny B} T$ if the refined kinetic energy is given by Eq.~\eqref{eq:modified kinetic energy}.

Our proposal for dealing with discretization effects on NVT simulations goes as follows.
It consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
Thus, we should adjust the Nos\'e-Hoover chain with the aim to obtain a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting, as described before.
The only required modification is that $\refined K$ should replace $K$ in Eq.~\eqref{eq:generalized force NVT}, which then becomes
\begin{equation}
\label{eq:thermostat_force}
G_1 = 2 \refined K - L k_\text{\tiny B} T.
\end{equation}

Also, the \textit{refined extended energy} preserved by this approach is now given by
\begin{equation}
\label{eq:this extended energy}
\widetilde{H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

One instant in which $\refined K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
Fortunately, only positions and orientations are required for the numerical differentiation.
This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time, as schematically illustrated in Fig.~\ref{fig:virtual}.
By incomplete we mean that the step is aborted once the positions and orientations have been updated.
\begin{figure}
\begin{tikzpicture}
\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
\node at (0.0,-0.2)[below]{$r_{t-2h}$};
\node at (2.2,-0.2)[below]{$r_{t-h}$};
\node at (4.4,-0.2)[below]{$r_{t}$};
\node at (6.6,-0.2)[below]{$r_{t+h}$};
\path[->,thick] (1) edge [bend left=45] (2);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
\path[dashed,->,thick] (1) edge [bend right=45] (0);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
\path[dashed,->,thick] (2) edge [bend left=45] (3);
\end{tikzpicture}
\caption{Schematic representation}
\label{fig:virtual}
\end{figure}

@@

\appendix
\numberwithin{equation}{section}

\section{Shadow Hamiltonian Approximation for Rigid Bodies}
\label{sec:rigid body shadow hamiltonian}

@@

As will become clear below, it is convenient to rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form
\begin{equation*}
K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} {\mt \Omega}(\vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation*}
where $\mt \Omega$ is a symmetric, block-diagonal matrix of size $7N \times 7N$ defined as
\begin{equation}
\label{eq:block-diagonal inverse mass tensor}
{\mt \Omega} = \left[\begin{array}{cc}
{\mt M}^{-1} & \mt 0 \\
\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
\end{array}\right].
\end{equation}
